Modelling the Co-evolution of Trade and Culture in Past Societies

### 2015-07-02 article.tex
new proof read of the article. miss a link to implmentation, and new results of simon

\paragraph{Production}

Each agent produces one good. The type of good produced by an agent $i$ is assigned to it at the beginning of the simulation, does not change through the simulation and is referred as $produced^i$. At each time step, each agent, produces a number of units (of its production good) equal to the number of goods, which ensures that enough is owned to be traded for other goods. Moreover, when an agent consumes its own production good, it does not impact its inventory.

@@

This function ensures that each good has the same weight in the final score (i.e.: managing to get only the right amount of a good with a high ``need'' value will not give a better score to the agent).
%\begin{figure}[htp]
%	\begin{center}
%		\includegraphics[width=7cm]{img/fitness.pdf}
%	\end{center}
%	\caption{The payoff depending on the quantity owned of a given good and the effective needs $n(g)$ for this good.}
%	\label{fig:fit}
%\end{figure}
%Figure~\ref{fig:fit} depicts the score of an agent, for different need values of one good, with regards to the quantity of the good possessed by the agent. We can see that even if the need for a good is small (for instance when $n(g) = 0.01$) an agent can have a high score as soon as it manages to get a quantity close to the need. This ensures that an agent needs to have for each of the good a quantity equal to the need and that no agent could reach a better score just by getting the right quantity for a good with a high need.
The complete score of an agent $i$ is termed $s^i$ and corresponds to the sum of the $s^i_j$. The agents will choose the model from whom the price vector should be copied among the agents that produce the same good and have the highest score. In practice, the worst two percent of the agents (in terms of score) will copy the prices of the best two percent. This selection process is detailed in algorithm~\ref{algo:selectionCulture}.

@@

If a trade is possible the two agents will exchange the agreed quantities. If the trade is not possible, the agent will continue to look at random partners for this good until either a partner is found or $TradeThreshold$ agents have been tried. At this point the agent will try to trade with agents producing another good. The process goes on until all goods have been tried. This trading process is described in algorithm~\ref{algo:trade}.

@@

\paragraph{Innovation}

In a trading environment it seems unlikely that a price will change radically to a very different value. Therefore, a new and more realistic mechanism is proposed. The innovation process, coined ``self referenced'', is still triggered with probability $\mu$ 
but modifies the previous price by adding or subtracting a small amount taken randomly from a uniform  distribution between $0 .. \beta$.

\paragraph{Expected outcome}

Based on the set of implementations presented and given the equations (\ref{eq:score}) and (\ref{eq:trade}), it is expected that the prices will converge to value allowing each agent to obtain quantities of resources exactly equivalent to the needs. The best possible price of all good satisfies the equations :
\begin{equation}
	\begin{cases}
		\frac{v^o_k}{v^o_g} = n_k \\
		v^o_g = n_g 
	\end{cases} =>v^o_k = n_k \times n_g, \quad \forall k \in Goods, \forall o \in Pop, g = produced^o, k \not= g 
\end{equation}

@@

The neutral model is tested through 15 experimental setups. The first six experimental setups are using 1 good, two population sizes (250 and 500 agents) and three values of $\mu$ (0.004,0.016 and 0.064). The remaining experimental setups are using 500 agents, 3 number of goods (3,6 and 9) and three values of $\mu$ (0.004,0.016 and 0.064). For each setup we have performed 100 runs of 10000 timestep each. The trading model is tested on an experimental setups using 3 goods, 500 agents and $\mu$ equal to 0.004. Again 100 runs of 10000 timesteps are performed. The complete set of parameters tested on both models is presented in table~\ref{tab:parameters}.

@@

\todo{complete table with final result (here it's just false exemple)}
\begin{table}[!h]
	\centering
	\begin{tabular}{cc|ccc}
		\multicolumn{2}{r}{}&\multicolumn{2}{c}{Our result}&\multicolumn{1}{c}{Bentley et. al. 2004}\\
			N&$\mu$ & $\alpha$ (sd) & $p$-value (sd - pr) &$\alpha$ (sd)\\\hline
		250	&0.004&1.53 (0.03)&0.58 (0.24 - .01)&1.54 (0.02)\\
			&0.016&1.57 (0.02)&0.35 (0.26 - .05)&1.57 (0.01)\\
			&0.064&1.66 (0.01)&0.0 (0.00 - 1)&1.67 (0.01)\\\hline
		500	&0.004&1.50 (0.02)&0.6 (0.05)&1.53 (0.03)\\
			&0.016&1.55 (0.03)&0.3 (0.05)&1.61 (0.04)\\
			&0.064&1.78 (0.08)&0.0 (0.00)&1.81 (0.10)\\
	\end{tabular}
	\caption{Mean $\alpha$ \& sd are calculate on 100 runs for our results, and 5 runs for Bentley et. al. 2004, pr is the percent of run for which the $p$-value<0.05 (we reject the hypothese that the distribution follow a power law).}
	\label{tab:mualpha}
\end{table}

@@

Nonetheless, for high value of $\mu$, previous works report that the distribution of variant follow a power law~\cite{bentley_random_2004}. This claim is based on the fact that the estimated $\alpha$ (estimation based on linear regression on the log-log curve) has a high correlation coefficient. Recent works have shown that the use of correlation coefficient should be avoided~\cite{clauset2009powerlawdistributionsinempiricaldat}. Following these recent findings, our results preclude us to assume that the distribution of variant when $\mu$ is up to $0.064$ follow a power law .

An additional series of experiments is done to analyse how the system reacts when multiple goods are present. The mean values of $\alpha$ and $p$-value have been analysed for 3 innovation rates (0.004, 0.016, 0.064), 4 number of goods (1, 3, 6, 9), and 250 agents. However, due to the simplicity of the results and the space restriction we only report the analysis of the results here. 
%\todo{complete table with final result (here it's just false exemple)}
%\begin{table}[!h]
%	\centering
%	\begin{tabular}{l|llll}
%		Innovation rate &\multicolumn{4}{c}{Number of goods}\\
%		      & 1   & 3   &  6 & 9  \\\hline
%		0.004 &1.53 (0.03)  &1.53 (0.03)&1.53 (0.03)&1.53 (0.03) \\
%		0.016 &1.57 (0.02)  &1.57 (0.02)&1.56 (0.017)&1.57 (0.02) \\
%		0.064 &1.66 (0.01)  &1.66 (0.01)&1.66 (0.01)&1.66 (0.01) \\\hline
%	\end{tabular}
%	\caption{Mean $\alpha$ and standard deviation for all innovation rate and different  of goods. The $\alpha$ given here is the mean $\alpha$ among all goods taken separately and computed over 100 runs with 250 agents.}
%	\label{tab:multiGoods}
%\end{table}
Visual analysis reveals that independently of the number of goods, the distributions obtained are exactly the same. Therefore it is no surprise that, for each innovation rate, the properties of the distribution of prices does not depend of the number of goods.

@@

In order to understand the effect of introducing trading mechanisms, we compare first the distribution of values obtained in the ``trading model'' against the values obtained in the ``neutral model''. The figure~\ref{fig:2setDi}.a) presents the results obtained from 100 runs for each model. All runs rely on the same experimental setup using 3 goods, 500 agents and $\mu$ equal to 0.004. In all following graph  a variant is one price for one given good. The distributions are first computed for each good independently and then averaged together.

@@

On figure~\ref{fig:2setDi}.a) it appears that the implementation of trade mechanisms leads to a distribution of prices departing from the neutral hypothesis. In more details, the proportion distribution has a plateau of common prices (a number of prices share similar and high proportion), which shows that, when trade is taken into account, the most common variants are more diverse.

@@

On figure~\ref{fig:2setDi}.b) it appears that the replacement of the innovation process leads to the creation of a distribution close to the one obtained with the neural model. On figure~\ref{fig:2setDi}.c) we observe a strong reduction in the size of the plateau and an important difference between the two distributions. This analysis highlights the importance of the \emph{innovation} process in the distribution of prices created by the trading model. This mechanism, by preventing the creation of totally random new price, leads to the creation of few similar prices obtaining high scores.

@@

We now study the results obtained in more detail by investigating the ability of the population to find the good price to perform exchanges between each other. This is done first by observing the score of all agents in each of the two different models. The figure~\ref{fig:scoreEvol} uses again the results obtained from 100 runs for each model where all runs rely on the same experimental setup using 3 goods, 500 agents and $\mu$ equal to 0.004. The figure is showing, as boxplots, the score computed thanks to equation (\ref{eq:score}) for all agents of all runs. The y-axis shows the score computed and the x-axis shows the timesteps. The left plot shows the results obtained in the ``cultural transmission'' model and the right plot shows the results obtained in the trading model.

@@

The previous figure showed the capacity for the trading model to increase its score but did not analyse the exact prices used. As explained in the section~\ref{sec:trade} we expect that the trade \emph{cultural transmission} and \emph{innovation} processes will produce a convergence toward a set of price for each good that will allow agents to exchange optimally the good they produce with the other goods. To verify this assumption we analyse the prices reached during the simulations. These are presented in figure~\ref{fig:ratioEvol} for the 100 runs relying the experimental setup using 3 goods, 500 agents and $\mu$ equal to 0.004. For all runs, all agents and at each iteration we compute the difference between the price used by the agent $V_g$ and the optimal price $\tilde{V}_g$. Record that the optimal price for a good is given in equation~\ref{eq:optimum}. All the measures performed are presented as boxplots where the results for 100 runs are condensed.

\begin{figure}[!h]
	\begin{center}
		\includegraphics[width=7cm]{img/ClearingPriceDistanceEvolutionForTrade-G3N500.pdf}
	\end{center}
	\caption{Evolution of the mean of the difference between the estimated value $v^i_g$ and the optimal value $\tilde{v}^i_g$ (as calculated with the equations~\ref{eq:optimum}) for a good $g$ and an agent $i$. As the optimal value $\tilde{v}^i_g$ depends on which good is produced by $i$, the mean of the difference between the estimated price and the optimal one is computed between all the agents that produce the same good. The figure represents this mean computed at each timestep for each goods, for each groups of agents and for 100 runs in a setup with 500 agents and 3 goods. }
	\label{fig:ratioEvol}
\end{figure}

@@

Additionally, more complex dynamics can be looked for. The trading mechanism can be naturally implemented in different ways, each reflecting a specific theory, and thus allowing their comparisons. The cultural transmission mechanism also can be modified by including for example the a crossover between the cultural trait of multiple agents. The trade network (which in this work can be interpreted as fully connected) can also be modified to study the effect of slow connections or the rupture of certain connections. The agent themselves could become more complex and be endowed with the ability to learn behaviours which would again produce more realistic simulations. Moreover, various populations can be envisioned, each with their distinctive characteristics regarding cultural transmission and trade.

>>>>

\paragraph{Production} Each agent produces one good. The type of good produced by an agent $i$ is assigned to it at the beginning of the simulation, does not change through the simulation, and is referred to as $produced^i$. At each time step, each agent, produces a number of units (of its production good) equal to the number of goods, which ensures that enough is owned to be traded for other goods. Moreover, when an agent consumes its own production good, it does not impact its inventory.

@@

This function ensures that each good has the same weight in the final score, i.e. managing to get only the right amount of a good with a high ``need'' value will not give a better score to the agent.
%\begin{figure}[htp]
%	\begin{center}
%		\includegraphics[width=7cm]{img/fitness.pdf}
%	\end{center}
%	\caption{The payoff depending on the quantity owned of a given good and the effective needs $n(g)$ for this good.}
%	\label{fig:fit}
%\end{figure}
%Figure~\ref{fig:fit} depicts the score of an agent, for different need values of one good, with regards to the quantity of the good possessed by the agent. We can see that even if the need for a good is small (for instance when $n(g) = 0.01$) an agent can have a high score as soon as it manages to get a quantity close to the need. This ensures that an agent needs to have for each of the good a quantity equal to the need and that no agent could reach a better score just by getting the right quantity for a good with a high need.
The complete score of an agent $i$ is termed $s^i$ and corresponds to the sum of the $s^i_j$. An agent will choose from whom the price vector should be copied among the agents that produce the same good and have the highest score. In practice, the worst two percent of the agents (in terms of score) will copy the prices of the best two percent. This selection process is detailed in algorithm~\ref{algo:selectionCulture}.

@@

If a trade is possible, the two agents will exchange the agreed quantities. If the trade is not possible, the agent will continue to look at random partners for this good until either a partner is found or $TradeThreshold$ agents have been tried. At this point the agent will try to trade with agents producing another good. The process goes on until all goods have been tried. This trading process is described in algorithm~\ref{algo:trade}.

@@

\paragraph{Innovation} In a trading environment it seems unlikely that a price will change radically to a very different value. Therefore, a new and more realistic mechanism is proposed. The innovation process, coined ``self referenced'', is still triggered with probability $\mu$ 
but modifies the previous price by adding or subtracting a small amount taken randomly from a uniform  distribution between $0 .. \beta$.

\paragraph{Expected outcome} Based on the set of implementations presented and given the equations (\ref{eq:score}) and (\ref{eq:trade}), it is expected that the prices will converge to value allowing each agent to obtain quantities of resources exactly equivalent to the needs. The best possible price of all good satisfies the equations :
\begin{equation}
	\begin{cases}
		\frac{v^o_k}{v^o_g} = n_k \\
		v^o_g = n_g 
	\end{cases} =>v^o_k = n_k \times n_g, \quad \forall k \in Goods, \forall o \in Pop, g = produced^o, k \not= g 
\end{equation}

@@

The neutral model is tested through 15 experimental setups. The first six experimental setups are using 1 good, two population sizes (250 and 500 agents) and three values of $\mu$ (0.004,0.016 and 0.064). The remaining experimental setups are using 500 agents, 3 number of goods (3,6 and 9) and three values of $\mu$ (0.004,0.016 and 0.064). For each setup, we have performed 100 runs of 10000 timestep each. The trading model is tested on an experimental setups using 3 goods, 500 agents and $\mu$ equal to 0.004. Again 100 runs of 10000 timesteps are performed. The complete set of parameters tested on both models is presented in table~\ref{tab:parameters}.

@@

\todo{complete table with final result (here it's just false exemple)}
\begin{table}[!h]
	\centering
	\begin{tabular}{cc|ccc}
		\multicolumn{2}{r}{}&\multicolumn{2}{c}{Our result}&\multicolumn{1}{c}{Bentley et. al. 2004}\\
			N&$\mu$ & $\alpha$ ($sd$) & $p$-value ($sd$ - $pr$) &$\alpha$ ($sd$)\\\hline
		250	&0.004&1.53 (0.03)&0.58 (0.24 - .01)&1.54 (0.02)\\
			&0.016&1.57 (0.02)&0.35 (0.26 - .05)&1.57 (0.01)\\
			&0.064&1.66 (0.01)&0.0 (0.00 - 1)&1.67 (0.01)\\\hline
		500	&0.004&1.50 (0.02)&0.6 (0.05)&1.53 (0.03)\\
			&0.016&1.55 (0.03)&0.3 (0.05)&1.61 (0.04)\\
			&0.064&1.78 (0.08)&0.0 (0.00)&1.81 (0.10)\\
	\end{tabular}
	\caption{Mean $\alpha$ \& $sd$ are calculated on 100 runs for our results, and 5 runs for Bentley et. al. 2004. $pr$ is the percentage of run for which the $p$-value is less than 0.05, i.e. the percentage of runs for which we rejected the null-hypotheses stating that the distribution follow a power law.}
	\label{tab:mualpha}
\end{table}

@@

Nonetheless, for high values of $\mu$, previous works report that the distribution of variant follow a power law~\cite{bentley_random_2004}. This claim is based on the fact that the estimated $\alpha$ (estimation based on linear regression on the log-log curve) has a high correlation coefficient. Recent works have shown that the use of correlation coefficient should be avoided~\cite{clauset2009powerlawdistributionsinempiricaldat}. Following these recent findings, our results preclude us to assume that the distribution of variant when $\mu$ is up to $0.064$ follow a power law .

An additional series of experiments is done to analyse how the system reacts when multiple goods are present. The mean values of $\alpha$ and $p$-value have been analysed for 3 innovation rates (0.004, 0.016, 0.064), 4 number of goods (1, 3, 6, 9), and 250 agents. However, due to the simplicity of the results and the space restriction we only report the analysis of the results here. 
%\todo{complete table with final result (here it's just false exemple)}
%\begin{table}[!h]
%	\centering
%	\begin{tabular}{l|llll}
%		Innovation rate &\multicolumn{4}{c}{Number of goods}\\
%		      & 1   & 3   &  6 & 9  \\\hline
%		0.004 &1.53 (0.03)  &1.53 (0.03)&1.53 (0.03)&1.53 (0.03) \\
%		0.016 &1.57 (0.02)  &1.57 (0.02)&1.56 (0.017)&1.57 (0.02) \\
%		0.064 &1.66 (0.01)  &1.66 (0.01)&1.66 (0.01)&1.66 (0.01) \\\hline
%	\end{tabular}
%	\caption{Mean $\alpha$ and standard deviation for all innovation rate and different  of goods. The $\alpha$ given here is the mean $\alpha$ among all goods taken separately and computed over 100 runs with 250 agents.}
%	\label{tab:multiGoods}
%\end{table}
Visual analysis reveals that independently of the number of goods, the distributions obtained are exactly the same. Therefore it is no surprise that, for each innovation rate, the properties of the distribution of prices does not depend on the number of goods.

@@

In order to understand the effect of introducing trading mechanisms, we compare first the distribution of values obtained in the ``trading model'' against the values obtained in the ``neutral model''. The figure~\ref{fig:2setDi}.a) presents the results obtained from 100 runs for each model. All runs rely on the same experimental setup using 3 goods, 500 agents and $\mu$ equal to 0.004. In all following graph  a variant is one price of one given good. The distributions are first computed for each good independently and then averaged together.

@@

On figure~\ref{fig:2setDi}.a) it appears that the implementation of trade mechanisms leads to a distribution of prices departing from the neutral hypothesis. In more details, the proportion distribution has a plateau of common prices (a number of prices share similar and high proportions), which shows that, when trade is taken into account, the most common variants are more diverse.

@@

On figure~\ref{fig:2setDi}.b) it appears that the replacement of the innovation process leads to the creation of a distribution close to the one obtained with the neural model. On figure~\ref{fig:2setDi}.c) we observe a strong reduction in the size of the plateau and an important difference between the two distributions. This analysis highlights the importance of the \emph{innovation} process of the trading model in the distribution of prices. This mechanism, by preventing the creation of totally random new price, promotes the creation of few similar prices.

@@

We now study the results obtained in more detail by investigating the ability of the population to find the good price to perform exchanges between each other. This is done first by observing the score of all agents in each of the two different models. The figure~\ref{fig:scoreEvol} uses again the results obtained from 100 runs for each model where all runs rely on the same experimental setup using 3 goods, 500 agents and $\mu$ equal to 0.004. The figure is showing, as boxplots, the score computed thanks to equation (\ref{eq:score}) for all agents of all runs. The y-axis shows the score computed and the x-axis shows the timesteps. The left plot shows the results obtained in the ``neutral model'' and the right plot shows the results obtained in the ``trading model''.

@@

The previous figure showed the capacity for the trading model to increase its score but did not analyse the exact prices used. As explained in the section~\ref{sec:trade} we expect that the trade \emph{cultural transmission} and \emph{innovation} processes will produce a convergence toward a set of price for each good that will allow agents to exchange optimally the good they produce with the other goods. To verify this assumption we analyse the prices reached during the simulations. These are presented in figure~\ref{fig:ratioEvol} for the 100 runs relying the experimental setup using 3 goods, 500 agents and $\mu$ equal to 0.004. For all runs, all agents and at each iteration we compute the difference between the price used by the agent $V_g$ and the optimal price $\tilde{V}_g$ (given by equation~\ref{eq:optimum}). The measures performed are presented as boxplots condensing the results for 100 runs.

\begin{figure}[!h]
	\begin{center}
		\includegraphics[width=7cm]{img/ClearingPriceDistanceEvolutionForTrade-G3N500.pdf}
	\end{center}
	\caption{Evolution of the mean of the difference between the estimated value $v^i_g$ and the optimal value $\tilde{v}^i_g$ (calculated with equation~\ref{eq:optimum}) for a good $g$ and an agent $i$. As the optimal value $\tilde{v}^i_g$ depends on which good is produced by $i$, the mean of the difference between the estimated price and the optimal one is computed between all the agents that produce the same good. The figure represents this mean computed at each timestep for each goods, for each groups of agents and for 100 runs in a setup with 500 agents and 3 goods. }
	\label{fig:ratioEvol}
\end{figure}

@@

Additionally, more complex dynamics can be looked for. The trading mechanism can be naturally implemented in different ways, each reflecting a specific theory, and thus allowing their comparisons. The cultural transmission mechanism also can be modified by including for example a crossover between the cultural trait of multiple agents. The trade network (which in this work can be interpreted as fully connected) can also be modified to study the effect of slow connections or the rupture of certain connections. The agent themselves could become more complex and be endowed with the ability to learn behaviours which would again produce more realistic simulations. Moreover, various populations can be envisioned, each with their distinctive characteristics regarding cultural transmission and trade.
